\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-13 article.tex
Cleanup

The implementation is written in OCaml for the most part, although there is some marginal use of Python to interact with the LLVM toolchain's debugger 'LLDB'.

@@

Instead of translating directly from a source language to target machine code, most compiler frontends emit IR which is intended as an abstraction over particular CPU implementations. % FIXME: er det korrekt? lol
It is by no means executable on either \(A\) or \(B\), but it is much closer to the instruction set executed on a CPU than the source language originally fed to the frontend.  This helps portability immensely, as a frontend emitting an IR no longer needs to specialize to a particular architecture, instead it can target as many as the backend supports.
Likewise, a compiler backend will support any frontend provided that they emit correct IR.

Because of this, compilation of high-level programming languages will often go no further than emitting IR, then leave the rest for a subsequent backend toolchain. This naturally leads to the LLVM toolchain, which is by far the most used compiler backend in practice. Most languages, if targeting native execution on a CPU, will have an LLVM implementation. This is true for C/C++ (Clang compiler \cite{clang}), D (LDC compiler \cite{dlang}), Swift (official Swift compiler \cite{swift}) and Rust (official \texttt{rustc} compiler \cite{rustc}). The use of LLVM allows for them to be compiled to completely unforeseen architectures, like web browsers with WebAssembly or even graphics cards with compute kernels \cite{backend}.

Conversely, managed languages such as Java and C\# target their own respective VMs, so they typically don't have much to gain from the LLVM toolchain. That said, some projects have attempted to implement LLVM in their translation of bytecode to native architectures,  such as LLILC which promised both \textit{just-in-time} (JIT) as well as \textit{ahead-of-time} (AOT) compilation \cite{dotnet}. Similarly, Kotlin (the spiritual successor of Java) that ordinarily targets the JVM has a native backend that compiles directly to machine code using LLVM, circumventing the need for a VM entirely \cite{kotlin}.
%Meanwhile interpreted languages like JavaScript eagerly attempt use LLVM for \textit{just-in-time} (JIT) compilation % FIXME: this is not true at all lol

%The modularization of the frontend and backend stages also helps immensely.
For example, the add function implemented in C as seen in Figure \ref{fig:addc} is compiled to the LLVM IR as seen in Figure \ref{fig:addll} after stripping optimization attributes with the \texttt{strip} utility. At this level of complexity they are largely equivalent, with the primary difference appearing to be syntactic. LLVM IR keeps the typed binary operations as well as the function construct with a return statement, although the variable names are discarded as IR is generally not for human interaction, except the function name, as this is used for linking with other object files.

\begin{figure}[H]
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{c}
int add(int a, int b) {
  int sum = a + b;
  return sum;
}
     \end{minted}
     \caption{Arithmetic function implemented in C}\label{fig:addc}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{llvm}
define i32 @add (i32 %0, i32 %1) {
 %3 = add i32 %1, %0
 ret i32 %3
}
     \end{minted}
     \caption{Stripped \texttt{clang -O1 -S -emit-llvm}}\label{fig:addll}
   \end{minipage}
   \begin{center}
   \begin{minipage}{0.83\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{gas}
	.text
	.globl	add               # -- Begin function add
add:                                    # @add
	movl	%esi, %eax
	addl	%edi, %eax
	retq
     \end{minted}
     \caption{Output of \mintinline{bash}{clang -S} without debug markers}\label{fig:addx86}
   \end{minipage}
   \end{center}
\end{figure}
When the IR is translated to x86 as seen in Figure \ref{fig:addx86}, the virtual variables \texttt{\%0}, \texttt{\%1} and \texttt{\%2} are assigned the registers \texttt{\%esi}, \texttt{\%edi} and \texttt{\%eax} respectively. This is presumably because of several optimizations, as \texttt{\%rdi} and \texttt{\%rsi} are used to pass the first and second parameters and the \texttt{\%rax} as the return value according to the System V AMD64 ABI \cite{sysv}, so by assigning the variables to those, several \texttt{mov}s are saved. Either way, the assignment is sound and the instruction selection is reasonably close to the  source code. There are other changes as well, as marking the target section with \texttt{.text} and the \texttt{.globl} annotation to export \texttt{add} in the symbol table which permits linking with other object files, but that isn't stricly related to register allocation.

@@

define i32 @mul (i32 %0, i32 %1) {
 %3 = icmp sgt i32 %0, 0
 br i1 %3, label %6, label %4
4:
 %5 = phi i32 [0, %2], [%9, %6]
 ret i32 %5
6:
 %7 = phi i32 [%10, %6], [0, %2]
 %8 = phi i32 [%9, %6], [0, %2]
 %9 = call i32 @add (i32 %8, i32 %1)
 %10 = add i32 %7, 1
 %11 = icmp eq i32 %10, %0
 br i1 %11, label %4, label %6
}
     \end{minted}
     \caption{Stripped \texttt{clang -O1 -S -emit-llvm}}\label{fig:mulll1}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{llvm}
define i32 @mul (i32 %0, i32 %1) {
 %3 = alloca i32
 %4 = alloca i32
 %5 = alloca i32
 %6 = alloca i32
 store i32 %0, i32* %3
 store i32 %1, i32* %4
 store i32 0, i32* %5
 store i32 0, i32* %6
 br label %7
7:
 %8 = load i32, i32* %6
 %9 = load i32, i32* %3
 %10 = icmp slt i32 %8, %9
 br i1 %10, label %11, label %18
11:
 %12 = load i32, i32* %5
 %13 = load i32, i32* %4
 %14 = call i32 @add (i32 %12, i32 %13)
 store i32 %14, i32* %5
 br label %15
15:
 %16 = load i32, i32* %6
 %17 = add i32 %16, 1
 store i32 %17, i32* %6
 br label %7
18:
 %19 = load i32, i32* %5
 ret i32 %19
}
     \end{minted}
     \caption{Stripped \texttt{clang -O0 -S -emit-llvm}}\label{fig:mulll0}
   \end{minipage}
\end{figure}
\noindent Another approach is that of Figure \ref{fig:mulll1}, which uses so-called 'phi nodes' instead.  These are a type of instruction that evaluates to a specified operand depending on which predecessor block was executed immediately prior. This circumvents the need for the same level of mutability as is achieved by reading/writing to memory, without violating the properties of SSA, by copying a value from the end of a predecessor to the beginning of the current block.
A Phi node represents a point in the program where the control flow merges, and it selects the appropriate version of a variable based on the path taken. This ensures that the data flow is well-defined and allows for easy analysis across different control flow paths.

@@

A continuous flow of instructions is referred to as a \textit{basic block}, defined as a sequence of instructions with no branches in or out except for the first instruction (referred to as a \textit{leader}, immediately following either the function entry or label within it) and the last (referred to as a \textit{terminator}, as it either terminates or transfers the flow of execution). %These are often referred to as terminators and can be thought of as an entirely different categoriy of instructions. % TODO: måske drop 'can be thought of'
In order to represent the order in which instructions are executed, a \textit{control flow graph} (CFG) is introduced. It is a directed graph whose edges represent transfer of control. The type of node varies over the source material, with CFGs of the Appel text \cite{tiger} constructed over individual instructions  and the Aho text \cite{dragon} over basic blocks.
% TODO: comment on precision
Either way, leaders will have a set of predecessors and terminators a set of successors associated with them.

Terminators have one or two successors in the case of branching or none at all in the case of function exit. % These can either be associated basic blocks or individual instructions that make it up.
Unconditional branching always transfers control to the block labelled, meaning only one successor follow, whereas conditional branching could transfer to either of the two, but because control flow analysis is not concerned with data, both are considered as possible successors. % TODO: rets

Successors of node \(n\) are denoted \( \mathit{succ}[n], \)
and while each node also has a set of predecessors associated, which consists of an  unbounded number of nodes from which control may be transferred, it isn't particularly useful in the following analysis.

@@

Building a graph is relatively straight forward, with the input for every function declared parsed as a tuple type named \mintinline{ocaml}{cfg}  of the form:
\begin{minted}{ocaml}
type cfg = (lbl option * block) * (lbl * block) list
\end{minted}
which consists of an optionally named entry block in the first part and a list of trailing blocks that are always named in the second. The reason for the first block to be optionally named is the fact that \textit{some} LLVM programs transfer control back to the point of entry, whereas every subsequent block needs to be named because branching can only target it by referring to its name.

@@

@@

\noindent Terminators are translated with branching instructions instead. One significant factor that needs to be accounted for is the phi nodes.
\begin{table}[H]
  \centering
\begin{tabular}{|>{\tt}l|>{\tt}l|l|} 
 \hline
 LLVM-{}- & x86 & Registers affected\\
 \hline
ret     &  retq & \{\}\\
br label \%l & jmp \%l & \{\}\\
br i1 \%c, label \%t, label \%f & cmpx \%c & \{\}\\
 & jz \%f & \\
 & jmp \%t & \\
 \hline
\end{tabular}
\caption{LLVM-{}- terminators to x86 translation table}\label{tab:trans-terms}
\end{table}

@@

Prime factorization of \lstinline!i32!.

@@

n & clang & simple 12 & simple 2 & briggs 12 & briggs 2 & linear & greedy 12 & greedy 0 \\

@@

268435399 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7783203125,0.9990234375,0.75} 1.11499x& \cellcolor[rgb]{0.77734375,0.9990234375,0.75} 1.11025x& \cellcolor[rgb]{0.7783203125,0.9990234375,0.75} 1.11595x& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.11943x& \cellcolor[rgb]{0.916015625,0.9990234375,0.75} 1.66718x& \cellcolor[rgb]{0.9990234375,0.9833984375,0.75} 2.47709x& \cellcolor[rgb]{0.9990234375,0.984375,0.75} 2.44535x\\
536870909 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64555x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64645x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.63974x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64095x& \cellcolor[rgb]{0.7607421875,0.9990234375,0.75} 1.04505x& \cellcolor[rgb]{0.9013671875,0.9990234375,0.75} 1.60979x& \cellcolor[rgb]{0.904296875,0.9990234375,0.75} 1.61997x\\
1073741789 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7958984375,0.9990234375,0.75} 1.18438x& \cellcolor[rgb]{0.7978515625,0.9990234375,0.75} 1.19584x& \cellcolor[rgb]{0.796875,0.9990234375,0.75} 1.19047x& \cellcolor[rgb]{0.7958984375,0.9990234375,0.75} 1.18472x& \cellcolor[rgb]{0.9990234375,0.9951171875,0.75} 2.09863x& \cellcolor[rgb]{0.9990234375,0.9560546875,0.75} 3.36121x& \cellcolor[rgb]{0.9990234375,0.9560546875,0.75} 3.37087x\\
2147483647 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8017578125,0.9990234375,0.75} 1.20872x& \cellcolor[rgb]{0.80078125,0.9990234375,0.75} 1.20665x& \cellcolor[rgb]{0.80078125,0.9990234375,0.75} 1.20405x& \cellcolor[rgb]{0.8017578125,0.9990234375,0.75} 1.20826x& \cellcolor[rgb]{0.9990234375,0.98828125,0.75} 2.32725x& \cellcolor[rgb]{0.9990234375,0.9453125,0.75} 3.71253x& \cellcolor[rgb]{0.9990234375,0.9453125,0.75} 3.71693x\\
4294967291 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.79296875,0.9990234375,0.75} 1.17597x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18070x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18067x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18159x& \cellcolor[rgb]{0.9990234375,0.986328125,0.75} 2.37710x& \cellcolor[rgb]{0.9990234375,0.9345703125,0.75} 4.04637x& \cellcolor[rgb]{0.9990234375,0.9345703125,0.75} 4.06492x\\
8589934583 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8203125,0.9990234375,0.75} 1.28534x& \cellcolor[rgb]{0.822265625,0.9990234375,0.75} 1.29317x& \cellcolor[rgb]{0.822265625,0.9990234375,0.75} 1.29145x& \cellcolor[rgb]{0.8212890625,0.9990234375,0.75} 1.28941x& \cellcolor[rgb]{0.9990234375,0.9755859375,0.75} 2.73679x& \cellcolor[rgb]{0.9990234375,0.9130859375,0.75} 4.74271x& \cellcolor[rgb]{0.9990234375,0.912109375,0.75} 4.76719x\\
17179869143 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.31790x& \cellcolor[rgb]{0.830078125,0.9990234375,0.75} 1.32297x& \cellcolor[rgb]{0.8310546875,0.9990234375,0.75} 1.32585x& \cellcolor[rgb]{0.830078125,0.9990234375,0.75} 1.32452x& \cellcolor[rgb]{0.9990234375,0.96875,0.75} 2.96961x& \cellcolor[rgb]{0.9990234375,0.900390625,0.75} 5.15949x& \cellcolor[rgb]{0.9990234375,0.9033203125,0.75} 5.07055x\\
34359738337 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.34905x& \cellcolor[rgb]{0.8369140625,0.9990234375,0.75} 1.34969x& \cellcolor[rgb]{0.8359375,0.9990234375,0.75} 1.34780x& \cellcolor[rgb]{0.845703125,0.9990234375,0.75} 1.38507x& \cellcolor[rgb]{0.9990234375,0.96484375,0.75} 3.08190x& \cellcolor[rgb]{0.9990234375,0.8935546875,0.75} 5.37886x& \cellcolor[rgb]{0.9990234375,0.892578125,0.75} 5.39434x\\

@@

268435399 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7783203125,0.9990234375,0.75} 1.11499x& \cellcolor[rgb]{0.77734375,0.9990234375,0.75} 1.11025x& \cellcolor[rgb]{0.7783203125,0.9990234375,0.75} 1.11595x& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.11943x& \cellcolor[rgb]{0.916015625,0.9990234375,0.75} 1.66718x& \cellcolor[rgb]{0.9990234375,0.9833984375,0.75} 2.47709x& \cellcolor[rgb]{0.9990234375,0.984375,0.75} 2.44535x\\
536870909 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64555x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64645x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.63974x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.64095x& \cellcolor[rgb]{0.7607421875,0.9990234375,0.75} 1.04505x& \cellcolor[rgb]{0.9013671875,0.9990234375,0.75} 1.60979x& \cellcolor[rgb]{0.904296875,0.9990234375,0.75} 1.61997x\\
1073741789 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.7958984375,0.9990234375,0.75} 1.18438x& \cellcolor[rgb]{0.7978515625,0.9990234375,0.75} 1.19584x& \cellcolor[rgb]{0.796875,0.9990234375,0.75} 1.19047x& \cellcolor[rgb]{0.7958984375,0.9990234375,0.75} 1.18472x& \cellcolor[rgb]{0.9990234375,0.9951171875,0.75} 2.09863x& \cellcolor[rgb]{0.9990234375,0.9560546875,0.75} 3.36121x& \cellcolor[rgb]{0.9990234375,0.9560546875,0.75} 3.37087x\\
2147483647 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1.00000x& \cellcolor[rgb]{0.8017578125,0.9990234375,0.75} 1.20872x& \cellcolor[rgb]{0.80078125,0.9990234375,0.75} 1.20665x& \cellcolor[rgb]{0.80078125,0.9990234375,0.75} 1.20405x& \cellcolor[rgb]{0.8017578125,0.9990234375,0.75} 1.20826x& \cellcolor[rgb]{0.9990234375,0.98828125,0.75} 2.32725x& \cellcolor[rgb]{0.9990234375,0.9453125,0.75} 3.71253x& \cellcolor[rgb]{0.9990234375,0.9453125,0.75} 3.71693x\\

>>>>

The implementation is written in OCaml for the most part, although  Python has been put to use to generate further tests paramterized over some value.% to interact with the LLVM toolchain's debugger 'LLDB'.

@@

Instead of translating directly from a source language to target machine code, most compiler frontends emit IR which is intended as an abstraction over CPU architectures. % FIXME: er det korrekt? lol
It is by no means executable on either \(A\) or \(B\), but it is much closer to the instruction set executed on a CPU than the source language originally fed to the frontend.  This helps portability immensely, as a frontend emitting an IR no longer needs to specialize to a particular architecture, instead it can target as many as the backend supports.
Likewise, a compiler backend will support any frontend provided that they emit correct IR.

Because of this, compilation of high-level programming languages will often go no further than emitting IR, then leave the rest for a subsequent backend implementation. This naturally leads to the LLVM toolchain, which is by far the most used compiler backend in practice. Most languages, if targeting native execution on a CPU, will have an LLVM implementation. This is true for C/C++ (Clang compiler \cite{clang}), D (LDC compiler \cite{dlang}), Swift (official Swift compiler \cite{swift}) and Rust (official \texttt{rustc} compiler \cite{rustc}) to name a few. In fact, the use of LLVM allows for them to be compiled to completely unforeseen and unintended targets, like web browsers with WebAssembly or even graphics cards with compute kernels \cite{backend}.

Conversely, managed languages such as Java and C\# target their own respective VMs, so they typically don't have much to gain from the LLVM toolchain. That said, some projects have attempted to implement LLVM in their translation of bytecode to native architectures,  such as LLILC which promised both \textit{just-in-time} (JIT) as well as \textit{ahead-of-time} (AOT) compilation \cite{dotnet}. Similarly, Kotlin (the spiritual successor to Java) that ordinarily targets the JVM has a native backend that compiles directly to machine code using LLVM, circumventing the need for a VM entirely \cite{kotlin}.
%Meanwhile interpreted languages like JavaScript eagerly attempt use LLVM for \textit{just-in-time} (JIT) compilation % FIXME: this is not true at all lol

%The modularization of the frontend and backend stages also helps immensely.
For example, the add function implemented in C as seen in Figure \ref{fig:addc} is compiled to the LLVM IR as seen in Figure \ref{fig:addll} after stripping optimization attributes with the \texttt{strip} utility. At this level of complexity they are largely equivalent, with the primary difference appearing to be syntactic. LLVM IR keeps the typed binary operations as well as the function construct with a return statement, although the variable names are discarded as IR is generally not for human interaction, except the function name, as this is used for linking with other object files where the symbol naming is used to look up offsets.

\begin{figure}[H]
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{c}
int add(int a, int b) {
  int sum = a + b;
  return sum;
}
     \end{minted}
     \caption{Arithmetic function implemented in C}\label{fig:addc}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{llvm}
define i32 @add (i32 %0, i32 %1) {
 %3 = add i32 %1, %0
 ret i32 %3
}
     \end{minted}
     \caption{Stripped \texttt{clang -O1 -S -emit-llvm}}\label{fig:addll}
   \end{minipage}
   \begin{center}
   \begin{minipage}{0.83\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{gas}
	.text
	.globl	add               # -- Begin function add
add:                                    # @add
	movl	%esi, %eax
	addl	%edi, %eax
	retq
     \end{minted}
     \caption{Output of \mintinline{bash}{clang -S} without debug markers}\label{fig:addx86}
   \end{minipage}
   \end{center}
\end{figure}
\noindent When the IR is translated to x86 as seen in Figure \ref{fig:addx86}, the virtual variables \texttt{\%0}, \texttt{\%1} and \texttt{\%2} are assigned the registers \texttt{\%esi}, \texttt{\%edi} and \texttt{\%eax} respectively. This is presumably because of several optimizations, as \texttt{\%rdi} and \texttt{\%rsi} are used to pass the first and second parameters and the \texttt{\%rax} as the return value according to the System V AMD64 ABI \cite{sysv}, so by assigning the variables to those, several \texttt{mov}s are saved. Either way, the assignment is sound and the instruction selection is reasonably close to the  source code. There are other changes as well, as marking the target section with \texttt{.text} and the \texttt{.globl} annotation to export \texttt{add} in the symbol table which permits linking with other object files, but that isn't stricly related to register allocation.

@@

define i32 @mul (i32 %0, i32 %1) {
 %3 = icmp sgt i32 %0, 0
 br i1 %3, label %6, label %4
4:
 %5 = phi i32 [0, %2], [%9, %6]
 ret i32 %5
6:
 %7 = phi i32 [%10, %6], [0, %2]
 %8 = phi i32 [%9, %6], [0, %2]
 %9 = call i32 @add (i32 %8, i32 %1)
 %10 = add i32 %7, 1
 %11 = icmp eq i32 %10, %0
 br i1 %11, label %4, label %6
}
     \end{minted}
     \caption{Stripped \texttt{clang -O1 -S -emit-llvm}}\label{fig:mulll1}
   \end{minipage}\hfill
   \begin{minipage}{0.48\textwidth}
     \centering
     \begin{minted}[bgcolor=LightGray]{llvm}
define i32 @mul (i32 %0, i32 %1) {
 %3 = alloca i32
 %4 = alloca i32
 %5 = alloca i32
 %6 = alloca i32
 store i32 %0, i32* %3
 store i32 %1, i32* %4
 store i32 0, i32* %5
 store i32 0, i32* %6
 br label %7
7:
 %8 = load i32, i32* %6
 %9 = load i32, i32* %3
 %10 = icmp slt i32 %8, %9
 br i1 %10, label %11, label %18
11:
 %12 = load i32, i32* %5
 %13 = load i32, i32* %4
 %14 = call i32 @add (i32 %12, i32 %13)
 store i32 %14, i32* %5
 br label %15
15:
 %16 = load i32, i32* %6
 %17 = add i32 %16, 1
 store i32 %17, i32* %6
 br label %7
18:
 %19 = load i32, i32* %5
 ret i32 %19
}
     \end{minted}
     \caption{Stripped \texttt{clang -O0 -S -emit-llvm}}\label{fig:mulll0}
   \end{minipage}
\end{figure}
\noindent Another approach is that of Figure \ref{fig:mulll1}, which uses so-called 'phi nodes' instead.  These are a type of instruction that evaluates to a specified operand depending on which predecessor block was executed immediately prior. This circumvents the need for the same level of mutability as is achieved by reading/writing to memory, without violating the properties of SSA, by copying a value from the end of a predecessor to the beginning of the current block.
A phi node represents a point in the program where the control flow merges, and it selects the appropriate version of a variable based on the path taken. This ensures that the data flow is well-defined and allows for easy analysis across different control flow paths.

@@

A continuous flow of instructions is referred to as a \textit{basic block}, defined as a sequence of instructions with no branches in or out except for the first instruction (referred to as a \textit{leader}, immediately following either the function entry or label within it) and the last (referred to as a \textit{terminator}, as it either terminates or transfers the flow of execution). %These are often referred to as terminators and can be thought of as an entirely different categoriy of instructions. % TODO: måske drop 'can be thought of'
In order to represent the order in which instructions are executed, a \textit{control flow graph} (CFG) is introduced. It is a directed graph whose edges represent transfer of control. The type of node varies over the source material, with CFGs of the Appel text \cite{tiger} constructed over individual instructions  and the Aho text \cite{dragon} over basic blocks.
% TODO: comment on precision
This project has opted to only use the Appel approach. One of the benefits is greater precision, in that several variables may be used in the same block and be assigned the same register as long as the two are not in interference within it. This wouldn't be possible with the Aho approach which completely discounts which parts of the block it is used in.
Either way, leaders will have a set of predecessors and terminators a set of successors associated with them.

Terminators have one or two successors in the case of branching or none at all in the case of function exit. % These can either be associated basic blocks or individual instructions that make it up.
Unconditional branching always transfers control to the block labelled, meaning only one successor follow, whereas conditional branching could transfer to either of the two, but because control flow analysis is not concerned with data, both are considered as possible successors. % TODO: rets 
Successors of node \(n\) are denoted \( \mathit{succ}[n], \)
and while each node also has a set of predecessors associated, which consists of an  unbounded number of nodes from which control may be transferred, it isn't particularly useful in the following analysis.

@@

Building a graph is relatively straight forward, with the input for every function declared parsed as a tuple type named \mintinline{ocaml}{cfg}  of the form:
\begin{minted}{ocaml}
type cfg = (lbl option * block) * (lbl * block) list
\end{minted}
which consists of an optionally named entry block in the first part and a list of trailing blocks that are always named in the second. The reason for the first block to be optionally named is the fact that \textit{some} LLVM programs transfer control back to the point of entry, whereas every subsequent block needs to be named because branching can only target it by referring to its name. Strictly speaking, this isn't how labels are treated in LLVM. Instead each block is always named, if not explicitly then by the next unused value from the same counter as temporaries \cite{llvm-functions}, but this was discovered too late in development to account for.

@@

\noindent As can be seen above, instructions always have one edge to the instruction/terminator immediately following, while terminators can have any number of edges (0 for \texttt{ret} and \texttt{unreachable}, 1-2 for \texttt{br} and an unbounded number for \texttt{switch}).

@@

% TODO explain terminators

@@

Integer factorization is another interesting algorithm. Generally considered to be NP, here are the results of factorizing known primes of inreasing bit length. For \(n=1024\).

@@

arg(s) & clang & simple 12 & simple 2 & briggs 12 & briggs 2 & linear 12 & linear 2 & greedy 12 & greedy 0 \\

@@

16777213 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 167130& \cellcolor[rgb]{0.759765625,0.9990234375,0.75} 1.04160x& \cellcolor[rgb]{0.7509765625,0.9990234375,0.75} 1.00651x& \cellcolor[rgb]{0.7626953125,0.9990234375,0.75} 1.05411x& \cellcolor[rgb]{0.7626953125,0.9990234375,0.75} 1.05372x& \cellcolor[rgb]{0.7607421875,0.9990234375,0.75} 1.04686x& \cellcolor[rgb]{0.958984375,0.9990234375,0.75} 1.84213x& \cellcolor[rgb]{0.9990234375,0.98828125,0.75} 2.33215x& \cellcolor[rgb]{0.9990234375,0.9873046875,0.75} 2.34511x\\
33554393 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 175409& \cellcolor[rgb]{0.7646484375,0.9990234375,0.75} 1.06219x& \cellcolor[rgb]{0.7646484375,0.9990234375,0.75} 1.06070x& \cellcolor[rgb]{0.765625,0.9990234375,0.75} 1.06415x& \cellcolor[rgb]{0.7646484375,0.9990234375,0.75} 1.06212x& \cellcolor[rgb]{0.7646484375,0.9990234375,0.75} 1.06039x& \cellcolor[rgb]{0.9990234375,0.9951171875,0.75} 2.12199x& \cellcolor[rgb]{0.9990234375,0.974609375,0.75} 2.78336x& \cellcolor[rgb]{0.9990234375,0.9736328125,0.75} 2.79956x\\
67108859 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 185779& \cellcolor[rgb]{0.7685546875,0.9990234375,0.75} 1.07588x& \cellcolor[rgb]{0.76953125,0.9990234375,0.75} 1.07923x& \cellcolor[rgb]{0.7705078125,0.9990234375,0.75} 1.08373x& \cellcolor[rgb]{0.76953125,0.9990234375,0.75} 1.08043x& \cellcolor[rgb]{0.76953125,0.9990234375,0.75} 1.07895x& \cellcolor[rgb]{0.9990234375,0.9833984375,0.75} 2.49571x& \cellcolor[rgb]{0.9990234375,0.955078125,0.75} 3.38089x& \cellcolor[rgb]{0.9990234375,0.955078125,0.75} 3.40121x\\
134217689 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 197302& \cellcolor[rgb]{0.7802734375,0.9990234375,0.75} 1.12173x& \cellcolor[rgb]{0.775390625,0.9990234375,0.75} 1.10225x& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.12120x& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.12080x& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.11794x& \cellcolor[rgb]{0.9990234375,0.9677734375,0.75} 3.00248x& \cellcolor[rgb]{0.9990234375,0.9306640625,0.75} 4.17628x& \cellcolor[rgb]{0.9990234375,0.9296875,0.75} 4.20718x\\
268435399 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 218313& \cellcolor[rgb]{0.7861328125,0.9990234375,0.75} 1.14571x& \cellcolor[rgb]{0.7861328125,0.9990234375,0.75} 1.14655x& \cellcolor[rgb]{0.7861328125,0.9990234375,0.75} 1.14531x& \cellcolor[rgb]{0.78515625,0.9990234375,0.75} 1.14376x& \cellcolor[rgb]{0.7861328125,0.9990234375,0.75} 1.14572x& \cellcolor[rgb]{0.9990234375,0.9501953125,0.75} 3.55232x& \cellcolor[rgb]{0.9990234375,0.9033203125,0.75} 5.06141x& \cellcolor[rgb]{0.9990234375,0.90234375,0.75} 5.08687x\\
536870909 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 245682& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18408x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18299x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18404x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18194x& \cellcolor[rgb]{0.794921875,0.9990234375,0.75} 1.18307x& \cellcolor[rgb]{0.9990234375,0.9296875,0.75} 4.20818x& \cellcolor[rgb]{0.9990234375,0.8720703125,0.75} 6.06747x& \cellcolor[rgb]{0.9990234375,0.8701171875,0.75} 6.13899x\\
1073741789 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 285913& \cellcolor[rgb]{0.8037109375,0.9990234375,0.75} 1.21667x& \cellcolor[rgb]{0.8037109375,0.9990234375,0.75} 1.21894x& \cellcolor[rgb]{0.8037109375,0.9990234375,0.75} 1.21755x& \cellcolor[rgb]{0.8037109375,0.9990234375,0.75} 1.21668x& \cellcolor[rgb]{0.8037109375,0.9990234375,0.75} 1.21706x& \cellcolor[rgb]{0.9990234375,0.908203125,0.75} 4.89647x& \cellcolor[rgb]{0.9990234375,0.8369140625,0.75} 7.20172x& \cellcolor[rgb]{0.9990234375,0.8359375,0.75} 7.23917x\\

@@

268435399 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 218996& \cellcolor[rgb]{0.7666015625,0.9990234375,0.75} 1.07020x& \cellcolor[rgb]{0.7763671875,0.9990234375,0.75} 1.10754x& \cellcolor[rgb]{0.767578125,0.9990234375,0.75} 1.07114x& \cellcolor[rgb]{0.7763671875,0.9990234375,0.75} 1.10950x& \cellcolor[rgb]{0.7744140625,0.9990234375,0.75} 1.09908x& \cellcolor[rgb]{0.90234375,0.9990234375,0.75} 1.61452x& \cellcolor[rgb]{0.9990234375,0.986328125,0.75} 2.38305x& \cellcolor[rgb]{0.9990234375,0.986328125,0.75} 2.38509x\\
536870909 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 247101& \cellcolor[rgb]{0.7705078125,0.9990234375,0.75} 1.08257x& \cellcolor[rgb]{0.7822265625,0.9990234375,0.75} 1.13159x& \cellcolor[rgb]{0.7705078125,0.9990234375,0.75} 1.08546x& \cellcolor[rgb]{0.7822265625,0.9990234375,0.75} 1.13164x& \cellcolor[rgb]{0.783203125,0.9990234375,0.75} 1.13364x& \cellcolor[rgb]{0.947265625,0.9990234375,0.75} 1.79438x& \cellcolor[rgb]{0.9990234375,0.9755859375,0.75} 2.73111x& \cellcolor[rgb]{0.9990234375,0.9755859375,0.75} 2.73643x\\
1073741789 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 285989& \cellcolor[rgb]{0.775390625,0.9990234375,0.75} 1.10503x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16197x& \cellcolor[rgb]{0.775390625,0.9990234375,0.75} 1.10359x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16233x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16392x& \cellcolor[rgb]{0.9912109375,0.9990234375,0.75} 1.97079x& \cellcolor[rgb]{0.9990234375,0.9638671875,0.75} 3.11293x& \cellcolor[rgb]{0.9990234375,0.9638671875,0.75} 3.12110x\\
2147483647 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 341802& \cellcolor[rgb]{0.779296875,0.9990234375,0.75} 1.12053x& \cellcolor[rgb]{0.7958984375,0.9990234375,0.75} 1.18807x& \cellcolor[rgb]{0.7802734375,0.9990234375,0.75} 1.12297x& \cellcolor[rgb]{0.7978515625,0.9990234375,0.75} 1.19256x& \cellcolor[rgb]{0.796875,0.9990234375,0.75} 1.19163x& \cellcolor[rgb]{0.9990234375,0.994140625,0.75} 2.14407x& \cellcolor[rgb]{0.9990234375,0.9521484375,0.75} 3.49973x& \cellcolor[rgb]{0.9990234375,0.951171875,0.75} 3.50847x\\
